---
title: "STA141A_Project"
output: pdf_document
date: "2025-05-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
```

# Joining and Cleaning Dataset

```{r}
# Predictors
demographics=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/Demographics.csv")
supplements=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/TotalSupplements.csv")
nutrients=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/TotalNutritionDay2.csv")
# Response variables
bloodpressure=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/BloodPressure.csv")
glucose=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/GlycatedHemoglobin.csv")
bodymeasurements=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/BodyMeasurements.csv")
totalchol=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/TotalCholesterol.csv")
hdl=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/HighDensityLipoprotein.csv")
```

```{r}
# Selecting relevant columns
demographics=demographics %>%
  select(SEQN,DMDBORN4,DMDEDUC2,DMDMARTZ,INDFMPIR,RIAGENDR,RIDAGEYR)
bloodpressure=bloodpressure %>%
  select(SEQN,BPXODI1,BPXODI2,BPXODI3,BPXOSY1,BPXOSY2,BPXOSY3)
bodymeasurements=bodymeasurements %>%
  select(SEQN,BMXBMI,BMXWAIST)
glucose=glucose %>%
  select(SEQN,LBXGH)
hdl=hdl %>%
  select(SEQN,LBDHDD)
totalchol=totalchol %>%
  select(SEQN,LBXTC)
nutrients=nutrients %>%
  select(SEQN,DR2_330Z,DR2TALCO,DR2TCARB,DR2TCHOL,DR2TFIBE,DR2TKCAL,DR2TMFAT,DR2TPFAT,DR2TPROT,DR2TSFAT,DR2TSODI,DR2TTFAT)
supplements=supplements %>%
  select(SEQN,DSQTCAFF,DSQTCALC,DSQTCHL,DSQTCOPP,DSQTFA,DSQTIODI,DSQTIRON,DSQTLYCO,DSQTLZ,DSQTMAGN,DSQTNIAC,DSQTPHOS,DSQTPOTA,DSQTSELE,DSQTVB1,DSQTVB2,DSQTVB12,DSQTVB6,DSQTVC,DSQTVD,DSQTVK,DSQTZINC)
```

```{r}
# Joining datasets
dataset=demographics %>%
  left_join(bloodpressure, by="SEQN") %>%
  left_join(bodymeasurements, by="SEQN") %>%
  left_join(glucose, by="SEQN") %>%
  left_join(hdl, by="SEQN") %>%
  left_join(totalchol, by="SEQN") %>%
  left_join(nutrients, by="SEQN") %>%
  left_join(supplements, by="SEQN")

# Making columns of average blood pressure, and column of Non-HDL cholesterol
dataset=dataset %>%
  mutate("AverageDiastolic"=rowMeans(select(.,BPXODI1,BPXODI2,BPXODI3),na.rm=T), "AverageSystolic"=rowMeans(select(.,BPXOSY1,BPXOSY2,BPXOSY3),na.rm=T)) %>%
  select(-BPXODI1,-BPXODI2,-BPXODI3,-BPXOSY1,-BPXOSY2,-BPXOSY3) %>%
  mutate("NonHDLChol"=ifelse(!is.na(LBXTC) & !is.na(LBDHDD), LBXTC-LBDHDD, NA))
```

```{r}
# Changing column names for ease of access
colnames(dataset)=c("SEQN","CountryofBirth", "HighestEducationLevel", "MaritalStatus","IncomeToPovertyRatio","Gender","AgeInYrs","BMI","WaistCircumference","HbA1c","HDLCholesterol","TotalBloodCholesterol","TotalWater","TotalAlcohol","TotalCarb","TotalCholesterolIntake","TotalFiberIntake","TotalCalories","TotalMUFAs","TotalPUFAs","TotalProtein","TotalSFAs","TotalSodium","TotalFat","CaffeineIntake","CalciumIntake","CholineIntake","CopperIntake","FolicAcidIntake","IodineIntake","IronIntake","LycopeneIntake","Lutein+ZeaxanthinIntake","MagnesiumIntake","NiacinIntake","PhosphateIntake","PotassiumIntake","SeleniumIntake","VitaminB1Intake","VitaminB2Intake","VitaminB12Intake","VitaminB6Intake","VitaminCIntake","VitaminDIntake","VitaminKIntake","ZincIntake","AverageDiastolic","AverageSystolic","NonHDLChol")
```

Country of birth is coded as: 1 (Born in the US), 2 (Born in other countries).

Highest education level is coded as: 1 (Less than 9th grade), 2 (9-11th grade), 3 (High school graduate or GED equivalent), 4 (Some college or Associate's degree), 5 (College graduate or above), 7 (Refused to answer, changed to NA), 9 (Don't know).

Marital status is coded as: 1 (Married/Living with partner), 2 (Widowed/Divorced/Separated), 3 (Never Married).

Gender is coded as: 1 (Male), 2 (Female).


#Edits made by Katherine (raw data sets)

#load in every data set
supplements <- read.csv("TotalSupplements.csv")
tnD1 <- read.csv("TotalNutritionDay1.csv") 
tnD2 <- read.csv("TotalNutritionDay2.csv")
cholesterol <- read.csv("TotalCholesterol.csv")
liver_ult <- read.csv("LiverUltrasound.csv")
insulin <- read.csv("Insulin.csv")
hsCRP <- read.csv("HsCRP.csv") #high sensitivity C-reactive protein
hdp <- read.csv("HighDensityLipoprotein.csv") 
glycated_hemoglobin <- read.csv("GlycatedHemoglobin.csv")
glucose <- read.csv("Glucose.csv")
demographics <- read.csv("Demographics.csv")
body_measurements <- read.csv("BodyMeasurements.csv")
bloodpressure <- read.csv("BloodPressure.csv")

data <- supplements |>
  full_join(tnD1, by = "SEQN") |>
  full_join(tnD2, by = "SEQN") |>
  full_join(cholesterol, by = "SEQN") |>
  full_join(liver_ult, by = "SEQN") |>
  full_join(insulin, by = "SEQN") |>
  full_join(hsCRP, by = "SEQN") |> 
  full_join(hdp, by = "SEQN") |>
  full_join(glycated_hemoglobin, by = "SEQN") |>
  full_join(glucose, by = "SEQN") |>
  full_join(demographics, by = "SEQN") |>
  full_join(body_measurements, by = "SEQN") |>
  full_join(bloodpressure, by = "SEQN")

glimpse(data)
summary(data)
head(data)

options(scipen = 999)

#count NAs
colSums(is.na(data))

#install.packages("Amelia")
library(Amelia)

#Visulaize missing values
missmap(data, main = "Missing Values Map")

library(ggplot2)

#Histogram
ggplot(data, aes(x = DR1TKCAL)) +
  geom_histogram(bins = 50, fill = "blue", color = "white") +
  labs(title = "Distribution of Total Energy Intake (Day 1)")

#Box plot for Body Mass Index (BMI)
ggplot(data, aes(y = BMXBMI)) +
  geom_boxplot(fill = "orange") +
  labs(title = "BMI Distribution")

ggplot(data, aes(x = factor(RIAGENDR))) +
  geom_bar(fill = "skyblue") +
  labs(title = "Gender Distribution", x = "Gender")

Bivariate Analysis
ggplot(data, aes(x = factor(RIAGENDR), y = DR1TKCAL)) +
  geom_boxplot() +
  labs(title = "Energy Intake by Gender", x = "Gender", y = "Energy Intake (kcal)")

ggplot(data, aes(x = DR1TKCAL, y = BMXBMI)) +
  geom_point(alpha = 0.4) + #alpha = transparency
  geom_smooth(method = "lm") + #regression line
  labs(title = "Calorie Intake vs BMI")


