---
title: "STA141A_Project"
output: pdf_document
date: "2025-05-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
```

# Joining and Cleaning Dataset

```{r}
# Predictors
demographics=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/Demographics.csv")
supplements=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/TotalSupplements.csv")
nutrients=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/TotalNutritionDay2.csv")
# Response variables
bloodpressure=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/BloodPressure.csv")
glucose=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/GlycatedHemoglobin.csv")
bodymeasurements=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/BodyMeasurements.csv")
totalchol=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/TotalCholesterol.csv")
hdl=read.csv("C:/Users/kjare/Downloads/drive-download-20250521T202410Z-1-001/HighDensityLipoprotein.csv")
```

```{r}
# Selecting relevant columns
demographics=demographics %>%
  select(SEQN,DMDBORN4,DMDEDUC2,DMDMARTZ,INDFMPIR,RIAGENDR,RIDAGEYR)
bloodpressure=bloodpressure %>%
  select(SEQN,BPXODI1,BPXODI2,BPXODI3,BPXOSY1,BPXOSY2,BPXOSY3)
bodymeasurements=bodymeasurements %>%
  select(SEQN,BMXBMI,BMXWAIST)
glucose=glucose %>%
  select(SEQN,LBXGH)
hdl=hdl %>%
  select(SEQN,LBDHDD)
totalchol=totalchol %>%
  select(SEQN,LBXTC)
nutrients=nutrients %>%
  select(SEQN,DR2_330Z,DR2TALCO,DR2TCARB,DR2TCHOL,DR2TFIBE,DR2TKCAL,DR2TMFAT,DR2TPFAT,DR2TPROT,DR2TSFAT,DR2TSODI,DR2TTFAT)
supplements=supplements %>%
  select(SEQN,DSQTCAFF,DSQTCALC,DSQTCHL,DSQTCOPP,DSQTFA,DSQTIODI,DSQTIRON,DSQTLYCO,DSQTLZ,DSQTMAGN,DSQTNIAC,DSQTPHOS,DSQTPOTA,DSQTSELE,DSQTVB1,DSQTVB2,DSQTVB12,DSQTVB6,DSQTVC,DSQTVD,DSQTVK,DSQTZINC)
```

```{r}
# Joining datasets
dataset=demographics %>%
  left_join(bloodpressure, by="SEQN") %>%
  left_join(bodymeasurements, by="SEQN") %>%
  left_join(glucose, by="SEQN") %>%
  left_join(hdl, by="SEQN") %>%
  left_join(totalchol, by="SEQN") %>%
  left_join(nutrients, by="SEQN") %>%
  left_join(supplements, by="SEQN")

# Making columns of average blood pressure, and column of Non-HDL cholesterol
dataset=dataset %>%
  mutate("AverageDiastolic"=rowMeans(select(.,BPXODI1,BPXODI2,BPXODI3),na.rm=T), "AverageSystolic"=rowMeans(select(.,BPXOSY1,BPXOSY2,BPXOSY3),na.rm=T)) %>%
  select(-BPXODI1,-BPXODI2,-BPXODI3,-BPXOSY1,-BPXOSY2,-BPXOSY3) %>%
  mutate("NonHDLChol"=ifelse(!is.na(LBXTC) & !is.na(LBDHDD), LBXTC-LBDHDD, NA))
```

```{r}
# Changing column names for ease of access
colnames(dataset)=c("SEQN","CountryofBirth", "HighestEducationLevel", "MaritalStatus","IncomeToPovertyRatio","Gender","AgeInYrs","BMI","WaistCircumference","HbA1c","HDLCholesterol","TotalBloodCholesterol","TotalWater","TotalAlcohol","TotalCarb","TotalCholesterolIntake","TotalFiberIntake","TotalCalories","TotalMUFAs","TotalPUFAs","TotalProtein","TotalSFAs","TotalSodium","TotalFat","CaffeineIntake","CalciumIntake","CholineIntake","CopperIntake","FolicAcidIntake","IodineIntake","IronIntake","LycopeneIntake","Lutein+ZeaxanthinIntake","MagnesiumIntake","NiacinIntake","PhosphateIntake","PotassiumIntake","SeleniumIntake","VitaminB1Intake","VitaminB2Intake","VitaminB12Intake","VitaminB6Intake","VitaminCIntake","VitaminDIntake","VitaminKIntake","ZincIntake","AverageDiastolic","AverageSystolic","NonHDLChol")
```

Country of birth is coded as: 1 (Born in the US), 2 (Born in other countries).

Highest education level is coded as: 1 (Less than 9th grade), 2 (9-11th grade), 3 (High school graduate or GED equivalent), 4 (Some college or Associate's degree), 5 (College graduate or above), 7 (Refused to answer, changed to NA), 9 (Don't know).

Marital status is coded as: 1 (Married/Living with partner), 2 (Widowed/Divorced/Separated), 3 (Never Married).

Gender is coded as: 1 (Male), 2 (Female).
